Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-06-08 writeup/rocha_etal.tex
Sec 3 check

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. The LLC2160 and LLC4320
simulations fully resolve internal tides and
other hydrostatic internal waves and other submesoscale processes at scales as
small as about 5 km.

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhance small-scale turbulence de-stratify the upper ocean, yielding mixed-layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed-layers as shallow as 40m.
The stratification and its seasonality is well-captured by both LLC simulations.
The excellent comparisons of model stratification to Argo climatology lends confidence
to the present analysis (supplemental material).

@@

\section{Statistics of surface lateral velocity tensor}

We present estimates of bulk quantities
associated with the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
and their analysis hint on the governing dynamics in the upper ocean
\citep{capet_etal2008a,shcherbina_etal2013}.

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions captures theseasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $1.5f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.4. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
and its local maximum and RMS are no larger than $0.5f$ and $0.2f$, respectively.
Indeed, the seasonal cycle in vorticity is very strong (figure~\ref{fig2}):
in both simulations, the RMS vertical vorticity and lateral strain rate are about twice as large in
late Winter/early Spring
than it is in late Summer/early Fall. Because the wintertime vorticity and strain
are dominated by the smallest scales in the flow, increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

The bulk of vertical vorticity and lateral strain rate is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km significantly reduces the RMS
vorticity and strain. The reduction in variance is dramatic ($>$80$\%$) in winter
yielding RMS vorticity and strain roughly consistent the ones diagnosed from
AVISO gridded geostrophic velocities (compare blue lines with black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergent part of the flow, however, add complexity
to the submesoscale-turbulence-centred scenario depicted by those invertigators.

 is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.9\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  lateral rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

@@

\section{Seasonality in governing submesoscale dynamics}

@@

\section{Projection on time scales}
To better understand the high-frequency/high-wavenumber flows  we calculate frequency
spectra of surface KE (figure \ref{fig4}). To obtain statistically meaningful results,
we average spectral estimates along a section at 165$^\circ$E.  There is about 50
independent spectral realizations.

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominated by inertia-gravity
waves. (Note that the inertial peak nearly merges with the
diurnal tide peak.) There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former flows
are more energetic in April and the latter in October.

Most of these high-frequency inertia-gravity waves project onto scales smaller
than 100 km. The surface KE spectra of 100-km-smoothed velocity field is significantly
suppressed in supra-inertial frequencies --- there are no statistically significant
changes at sub-inertial frequencies (see dashed lines in figure \ref{fig4}).

>>>>

A key aspect of the LLC simulations is that there were forced by
the 16 most significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhance small-scale turbulence de-stratify the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
The stratification and its seasonality is well-captured by both LLC simulations:
the skillful comparisons of model stratification to Argo climatology lends confidence
to the present analysis (supplemental material).

@@

\section{Statistics of surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
and their analysis hint on the governing dynamics in the upper ocean
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early Spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.4. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
and its local maximum and RMS are no larger than $0.5f$ and $0.2f$, respectively.
Indeed, the seasonal cycle of vorticity and rate of strain is very strong (figure~\ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late Winter/early Spring
than in late Summer/early Fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow, increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.8\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  lateral rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

The bulk of vertical vorticity and lateral strain rate is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent the ones diagnosed from
AVISO gridded geostrophic velocities (compare blue lines with black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}a; RMS divergence $\sim$0.22 f in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late Summer/early Fall (RMS divergence $\sim$0.22 f the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolve
smaller-scale submesoscale flows, the
winter peak in divergence is nearly as strong the summertime's. Of course, part of
these submesoscale fronts and eddies evolve relatively fast and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c strongly suggest that the
summertime submesoscale surface variability stems from different dynamics than
then wintertime variability. To better understand these differences, we calculate
joint probability distributions...

@@

@@

% \section{Projection on time scales}
% To better understand the high-frequency/high-wavenumber flows  we calculate frequency
% spectra of surface KE (figure \ref{fig4}). To obtain statistically meaningful results,
% we average spectral estimates along a section at 165$^\circ$E.  There is about 50
% independent spectral realizations.
%
% The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
% The high-frequency flows are split near-inertial flows, tidal signals and their
% higher harmonics. Hence these high-frequency flows are dominated by inertia-gravity
% waves. (Note that the inertial peak nearly merges with the
% diurnal tide peak.) There is statistically significant seasonality. The seasonal
% cycle of the sub-inertial motions (at least for periods between 5-20 days)
% is out-of-phase with the seasonal cycle of supra-inertial flows: the former flows
% are more energetic in April and the latter in October.
%
% Most of these high-frequency inertia-gravity waves project onto scales smaller
% than 100 km. The surface KE spectra of 100-km-smoothed velocity field is significantly
% suppressed in supra-inertial frequencies --- there are no statistically significant
% changes at sub-inertial frequencies (see dashed lines in figure \ref{fig4}).
